A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Shorten the statement of trichotomy

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  Exactly one of the following statements holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.

@@

Now we proceed with the definition of mode-correctness.
As we have outlined at the beginning, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Hence, we instead choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align*}
    \synvar(\cdot)                                  & = \emptyset  \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align*}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

@@

>>>>

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  Exactly one of the following holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
    \item $\isTerm{t}$ is synthesising but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$ for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.

@@

Now we proceed with the definition of mode-correctness.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Instead, we choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align*}
    \synvar(\cdot)                                  & = \emptyset  \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align*}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

@@

\begin{proof}
